Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2024-03-06 revision/review_response.qmd

*We thank all three reviewers for providing thorough and constructive comments that have helped to improve the manuscript. Our responses to each of these comments are provided below. Note that line numbers refer to those in the original manuscript.*

### Reviewer 1

This manuscript describes an advancement in the computation of aquatic estuarine metabolism (i.e., primary production and reparation) using open-water dissolved oxygen (DO) time series. The authors adapted a Bayesian model originally developed for non-tidal, lotic systems (i.e., BASE) to 1) now work with time series that have missing dissolved oxygen observations and 2) can accommodate longer time periods for metabolism estimates. The latter could produce non- real estimates of metabolism in earlier methods while the former is achieved through linear interpolation during pre-processing and estimate removal by a user defined threshold during post-processing. The authors describe in sufficient detail how to operationalize the model and provide a comparison with a historically more standard arithmetic-based model that has been used in earlier studies of estuarine metabolism. Furthermore, the authors provide an R package “EBASE” to allow others to estimate estuarine metabolism using this method advancement; a package I am sure the coastal community will find useful. Overall, this paper was well written and the advancement significant. This paper should be considered for publication in L&O Methods following incorporation of the comments below.

*Response: Thank you for providing these thorough comments on our manuscript.  Please see our responses below.*

@@

There were two studies cited in the introduction (Line 98) as having used Bayesian techniques to estimate aquatic metabolism in estuaries. Please elaborate on how the method described in this study is consistent with and/or different from those in the cited studies.

*Response: We agree this clarification is needed.  The first study [@ciavatta2008] actually did not use a Bayesian approach, rather it provided simultaneous estimation of the metabolic parameters using a conventional empirical model framework.  This reference to a Bayesian approach was in error and we have removed this citation from the introduction. @tassone2019 does provide a good example for comparison with our work and we elaborate on this in the revised text on line 105: "@tassone2019 adapted a Bayesian approach developed for streams [@appling2018] in an estuarine setting to estimate metabolic rates in the James River.  Although this approach demonstrated the utility of such an approach in estuaries, much of the core model structure developed for streams was unchanged and may not have been fully descriptive of fundamental processes influencing metabolism in open-water environments."*

*We then continue the justification for our use of BASE, adding the following:*

*Line 108: "The underlying model components can also easily be updated given the relatively simple format of the approach."*

*Line 115: "We focus on BASE given the relative simplicity and, therefore, ease of application compared to other stream-based Bayesian approaches [e.g., @appling2018]."*

*The streamMetabolizer package is a great contribution to the field, although our implicit goal of this work was to modify a readily accessible open source approach such as BASE for our needs.  The framework was easily extendable for our chosen gas-exchange method [@Wanninkhof14] and modification of the software functionality (i.e., user-chosen optimization period and dealing with data gaps).*

@@

The authors describe pitfalls in their model, particularly when it comes to estimating the gas transfer velocity. Additionally, the authors provide evidence that sometimes the right answer (i.e., modeled DO time series, lines 349-352) can be achieved through incorrect processes. Similar issues have been described and addressed in the stream metabolism literature which was not cited or discussed in this study, please see Appling et al. 2018 - https://doi.org/10.1002/2017JG004140). Consider adding to the discussion how EBASE might incorporate these advancements to address these issues.

*Response: Thank you for pointing out this reference.  We have included an explicit mention of equifinality and a citation for @appling2018 on line 352.  Also please see our addition to the discussion on line 490.  This is described in full in response to the general comments from Reviewer 3.*

@@

Line 96: Add the word statistical between ‘Bayesian’ and ‘techniques’ (i.e., Bayesian statistical techniques…)

*Response: Added.*

Line 98: These two cited studies – Ciavatta et al. 2008; Tassone and Bukaveckas 2019 - have used Bayesian approaches to estimate estuarine metabolism; what makes them limited? How is the approach described in this manuscript different from these two earlier studies? Have those earlier two studies informed the model selection/development described in this manuscript?

*Response: Please see our response to your general comment above.*

@@

Line 324-327: I'm not sure if I follow correctly - are you seeking daily metabolic estimates? If so, are the 7 & 30 day optimizations using a forward looking moving window approach that advances by one-day (similar to a rolling average)?

*Response: The text was edited for clarity: "Results from optimization periods of 1, 7, and 30 days were compared to the synthetic one-year time series ($b$ was fixed for the sensitivity analysis) based on daily averages for both the synthetic and EBASE results. For the latter, parameters that are only returned as individual values for each optimization period ($a$, $R$, and $b$) were assigned as daily values for comparability of summary statistics where $n$ is the same across analyses."*

Line 473-477: It is stated that “Bayesian…methods have not been rigorously explored for estimating metabolic rates in estuaries” however, I point back to the two cited studies in the introduction (line 98) that did use Bayesian techniques to estimate estuarine metabolism. Consider rewording this sentence and/or expand on why those earlier studies are not considered rigorous.

*Response: Please see our response to your general comment above.  Although @tassone2019 does provide another useful example, we still feel that these methods have not been thoroughly explored in estuaries.*

@@

*Response: Apologies for this confusion. Eq. 3 is not used in EBASE, rather it is discussed as the gas transfer used in BASE that is inappropriate for our application, justifying use of the gas exchange model in @Wanninkhof14. Line 166 was revised to make this more clear: "Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE (Equation 2)..."*

Diving into eq 3, higher water levels usually indicate higher flows and thus higher gas transfer velocity (Kw). I’m not exactly sure how temperature plays into this equation, and if it is appropriate for an estuary. It seems to suggest warm periods within a day have higher exchange rates, due to differences in kinetics at the water surface. However warmer temperatures may also reflect increased stratification and less vertical mixing (small lakes). There may be something to clean from changes in temperature that drive changes in turbulence, but simply using warmer = more exchange seems like an oversimplification. Eq 3 also includes a term related to water depth (H). In flowing waters this equation suggests that Kw increases with water level, likely a result of high-water levels = higher velocities = higher turbulence = higher kw. In an estuary, I’m not sure this relationship truly holds as water depth is not as good of a surrogate for velocity. Highest water level may be at slack tide and thus minimal water movements. It could be that change in water depth (assuming the tidal currents are strong relative to the river inputs) would be a better term to reflect instantaneous discharge. Or possibly flow could be included as a possible input parameter. But I realize this term is not typically available.

*Response: We have noted an error in our description of Equation 3 that includes the depth-dependency.  As noted by Reviewer 3, @Grace15 do not include a depth-dependency in this equation. We have since removed the dependency and revised the text accordingly, emphasizing the importance of wind for the transfer velocity in open-water systems.  Regarding temperature, we feel justified in its inclusion for gas exchange given the previously noted relationship with the Schmidt number, a known factor to be important in open-water systems.  Also see our response to your comment on line 164.*

@@

All told, perhaps the issue with poorly constraining b is related to assumptions in the equations used to model gas exchange in an estuary. The authors do a good job explaining that the model parameter b was difficult to constrain in the example dataset. This is understandable, and my comments are mainly my mental exploration of ways that might improve that aspect of the model. I’m guessing the authors have thought extensively about this problem, and I’m mainly offering ideas. Perhaps the authors could include some other ideas for why the b parameter is difficult to constrain? Or explore additional datasets to understand when/where it can be estimated more accurately? For example, if the model was applied to an artificial dataset from a stream, lake, or other type of estuary does it perform better? If so, what is different about the system, which might help with tuning the model? Would any of these help understand the parameterization/limitation of the current model, and or identify systems it is most appropriate?

*Response: Yes, we agree this issue is perplexing and worth exploring further.  We have mentioned in the original draft how this could be explored further, e.g., evaluation in other locations where gas exchange may be a more dominant component of the metabolic model (line 490).  We have also added a brief description of @appling2018 that confront this issue in stream metabolic models.  Please see our response to the general comments from Reviewer 3.*

@@

In particular, the focus on the importance of accurately capturing the physical variables is a strength of the paper. It is made clear from the outcomes of the synthetic dataset analysis that it is better to use a literature estimate of the parameter b, a constant in the equation that governs air-water exchange, then to let the model estimate inaccurate values for b. The discussion around this issue was appropriate, but I wonder if some more very clear advice could be provided to a potential user that wants to use this package to answer questions about ecosystem metabolism in estuaries but does not have the aim of further refining the modelling approaches. Could they use the estimate from Wanninkhof (2014) as was used in subsequent analyses here? Would an appropriate sensitivity analysis be to vary this estimate by 20% (as mentioned L224) to check if their P and R results are highly sensitive to this parameter? Or is further testing in a range of estuaries required before this can be applied?

*Response: Thank you for these suggestions.  We agree an explicit description of how the approach can be applied in light of uncertainties in the $b$ parameter is needed.  Some discussion is already included (e.g., line 488).  We have also added text describing how @appling2018 addressed the issue, although this suggestions would be further model development.  Revisions were made to the first paragraph of the discussion.*

*Line 490: "For application, users of EBASE may simply use a fixed value for $b$ (e.g., 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$) or perform a simple sensitivity analysis by varying $b$ +/- a small percentage to determine how $P$ and $R$ may change.  Further model development may consider alternative approaches for dealing with equifinality, i.e., when the same modeled DO time series can be returned with different estimates of $P$, $R$, and $D$.  @appling2018 provide a useful example for a stream metabolic model that addresses equifinality by allowing model parameters to vary with key drivers at sub-daily scales (e.g., gas exchange versus discharge) and avoiding assumptions that model residuals are uncorrelated."*

@@

L269: Should it be “availability” rather than “variability”? of course, a more variable pool of organic matter could also be correct.

*Response: Changed to "availability", this seems more correct.*

L352: The “right answer for wrong reasons” is interesting. It could be clearer here that even though DO is estimated accurately, the P and R estimates will be inaccurate in this scenario.

*Response: Sentence was revised as follows: "...the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ (Equation 1). This issue, or equifinality, has been encountered in stream metabolic models [@appling2018]."*

### 2024-03-06 manu-draft.qmd
revisions done but need to proof and add cran

A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Bayesian statistical techniques have only been explored in limited estuarine applications despite their potential to address the above [@ciavatta2008; @tassone2019]. Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016]. These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of parameters related to gas exchange, P, and R (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange. A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15]. BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations. As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, metabolic estimates from EBASE are compared with those of the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series, which had a median NSE across all optimization periods and priors of `r prmall[['DO_mod']]`. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ ([@eq-ebase]). After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`. This result makes sense, given that differences in $D$ across all optimization periods and priors are due solely to difference in $DO$. Conversely, the $a$ parameter was reproduced the most poorly, with a median NSE of `r prmall[['a']]`.

@@

The above analyses demonstrated that EBASE can successfully recover known metabolic parameters from a synthetic time series and provides robust estimates on real data covering a year of observations. The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications. These needs were addressed primarily using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit. This framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13]. The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries. Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty. However, initial analyses suggested an inability of the model to estimate stable values for the $b$ parameter that influence gas exchange. This parameter was fixed for the sensitivity analyses and we were unable to evaluate the ability of EBASE to recover it from the synthetic time series. Future work should focus on understanding the sensitivity of EBASE to estimating $b$ where the dominance of the relative processes that manifest signals in the dissolved oxygen time series could vary. For example, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data may provide insufficient information for EBASE to estimate $b$. This aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries of the Gulf of Mexico [@Murrell18]. In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with expected larger contributions of gas exchange expressed by dissolved oxygen.

>>>>

A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Bayesian statistical techniques have only been explored in limited estuarine applications despite their potential to address the above [e.g., @tassone2019]. Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016]. These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of parameters related to gas exchange, P, and R (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange. @tassone2019 adapted a Bayesian approach developed for streams [@appling2018] in an estuarine setting to estimate metabolic rates in the James River. Although this approach demonstrated the utility of such an approach in estuaries, much of the core model structure developed for streams was unchanged and may not have been fully descriptive of fundamental processes influencing metabolism in open-water environments. Another potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15]. BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations. The underlying model components can also easily be updated given the relatively simple format of the approach. As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. We focus on BASE given the relative simplicity and, therefore, ease of application compared to other stream-based Bayesian approaches [e.g., @appling2018]. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, metabolic estimates from EBASE are compared with those of the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study. EBASE is available as an open-source software package created with the R statistical programming language \[@Beck23; @RCT23\] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series, which had a median NSE across all optimization periods and priors of `r prmall[['DO_mod']]`. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ ([@eq-ebase]). This issue, or equifinality, has been encountered in stream metabolic models [@appling2018]. After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`. This result makes sense, given that differences in $D$ across all optimization periods and priors are due solely to difference in $DO$. Conversely, the $a$ parameter was reproduced the most poorly, with a median NSE of `r prmall[['a']]`.

@@

The above analyses demonstrated that EBASE can successfully recover known metabolic parameters from a synthetic time series and provides robust estimates on real data covering a year of observations. The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications. These needs were addressed primarily using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit. This framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13]. The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries.

Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty. However, initial analyses suggested an inability of the model to estimate stable values for the $b$ parameter that influence gas exchange. This parameter was fixed for the sensitivity analyses and we were unable to evaluate the ability of EBASE to recover it from the synthetic time series. Future work should focus on understanding the sensitivity of EBASE to estimating $b$ where the dominance of the relative processes that manifest signals in the dissolved oxygen time series could vary. For example, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data may provide insufficient information for EBASE to estimate $b$. This aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries of the Gulf of Mexico [@Murrell18]. In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with expected larger contributions of gas exchange expressed by dissolved oxygen. For application, users of EBASE may simply use a fixed value for $b$ (e.g., 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$) or perform a simple sensitivity analysis by varying $b$ +/- a small percentage to determine how $P$ and $R$ may change. Further model development may consider alternative approaches for dealing with equifinality, i.e., when the same modeled DO time series can be returned with different estimates of $P$, $R$, and $D$. @appling2018 provide a useful example for a stream metabolic model that addresses equifinality by allowing model parameters to vary with key drivers at sub-daily scales (e.g., gas exchange versus discharge) and avoiding assumptions that model residuals are uncorrelated.
